Metropolitan Statistical Area Evaluation

### 2022-11-11 src/demo_data_wrangling.Rmd
updating population sixe, including normalized population and population estimate

# create __ 
potential_MSAs <- msa_pop_2021_data %>% 
  filter(pop_percent_change > 0.5 ) %>% 
  mutate(pop_percent_change20_21 = round(pop_percent_change, digits = 3)) %>% 
  filter (!state %in% c("CA", "HI", "AK")) %>% 
  #separate(col = "state", into = c("state1", "state2", "state3"),    sep = "-") %>% 
  filter(!cbsa %in% c("10580", "10900", "12420", "12940", "14260", "15980", "18580", "19100", "19660", "19740", "23540", "25420", "26420", "27260", "29820", "34940", "36260", "36740", "38060", "39100", "39300", "39340", "41620", "45220", "45300", "46060", "46140", "39900")) %>%  # remove MSAs as instructed by Owen 
  mutate(normalized_pop2021 = (popestimate2021-min(popestimate2021))/ (max(popestimate2021)-min(popestimate2021))) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021)

@@

# join population growth data set to household median income
income_potential_MSAs <- MSA_income_data %>% # rename row names
  select(cbsa, median_income) %>% 
  inner_join(potential_MSAs) %>% 
  select(cbsa, MSA, state, median_income, pop_percent_change20_21, normalized_pop2021) #reorder columns
#select(cbsa, MSA, state, pop_percent_change20_21, median_income, income_moe_USD)

@@

# data wrangling, only selecting data for potential MSAs
MSA_employment <- unemployment_data %>% 
  filter(year >= 2020) %>% 
  full_join(income_potential_MSAs) %>% 
  drop_na() %>% 
  select(cbsa, MSA, year, month, employment_rate)

avg_emp_yoy <- MSA_employment %>% 
  group_by(MSA, year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  mutate(avg_emp_YoY = (avg_employment_rate-lag(avg_employment_rate))/lag(avg_employment_rate)*100) %>% 
  drop_na()

# calculate YoY growth 
MSA_employment_yoy <- MSA_employment %>% 
  filter(year >= 2020) %>% 
  filter(month == "01") %>% 
  arrange(MSA, month) %>% 
  group_by(MSA) %>% 
  mutate(YoY = (employment_rate-lag(employment_rate))/lag(employment_rate)*100) %>% 
  mutate(employment_YoY = round(YoY, digits = 3)) %>% 
  mutate(emp_pos = employment_YoY >= 0) %>%# function to create column for yoy value, true = positive YoY
  inner_join(avg_emp_yoy) %>% 
  filter(year == "2022") %>% 
  select(!c(YoY, month, year))

@@

potential_MSAs <- potential_MSAs %>%

@@

```{r eval = FALSE}
# average employment from 2000 to 2022
employment_summary_since2000 <- MSA_employment %>% 
  group_by(MSA) %>% 
  filter(year >= 2000) %>% 
  summarize(
    avg_employment = mean(employment_rate),
    sd_employment = sd(employment_rate)
  )

```{r employment visuals, fig.align = "center", fig.show="hold", fig.height= 7, fig.width=10}
# visual of MSA employment rates, shows major drop during covid
ggplot(MSA_employment_yoy, aes(x = year, y = employment_rate, color = MSA)) +
  geom_line() +
  xlim(2021, 2022) +
  theme(legend.position = "none") +

@@

# year over year change in employment rate, split up by year
ggplot(MSA_employment_yoy, aes(x = employment_YoY, y = reorder(MSA, employment_YoY), fill = emp_pos)) +
  geom_col(position = "identity", color = "black", size = 0.25) +
  scale_fill_manual(values = c("red", "darkolivegreen3"), guide = FALSE) +
  facet_grid(~year) +
  labs(x = "Year-over_year change in employment rate",
       y = "")

@@

```
```{r payroll visual, fig.height = 7, eval = FALSE}
ggplot(MSA_payroll, aes(x = payroll_change, y = reorder(MSA, payroll_change), fill = payroll_pos)) +
  geom_col(position = "identity", color = "black", size = 0.25) +
  scale_fill_manual(values = c("red", "darkolivegreen3"), guide = FALSE) +
  labs(x = "percent change",
       y = "",
       title = "Quarterly change in employee payroll from 2021 to 2022") +
  facet_grid(~year)

@@

MSA_median_rent2022_data <- read_csv(here("data", "raw", "2022MSA_median_rent.csv")) %>% 
  separate(col = cbsasub22,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = as.double(cbsa)) %>% 
  mutate(median_rent2022 = as.numeric(rent_50_0)) %>% 
  select(cbsa, median_rent2022)

potential_MSAs <- potential_MSAs %>% 
  right_join(MSA_median_rent2022_data) %>% 
  drop_na()

@@

```{r house price data}
metro_house_cost_data <- read_csv(here("data","raw", "metro_home_costs", "metro_home_prices.csv")) %>% 
  row_to_names(3) %>% 
  clean_names() %>% 
  separate(col = "metropolitan_area",
           into = c("cbsa", 'metro_area'),
           sep = 5) %>% 
  mutate(cbsa = as.numeric(cbsa)) %>% 
  mutate(x2020 = as.numeric(x2020))

metro_house_prices <- MSA_house_income %>%
  left_join(metro_house_cost_data)  %>% 
  rename(house_price_2020 = x2020) %>% 
  rename(house_price_2021 = x2021_r) %>% 
  rename(houseprice_change21_22 = q1_q1) %>% 
  select(!c(metro_area, x2019, na)) %>% 
  # select(cbsa, houseprice_change21_22) %>% 
  separate(col = "houseprice_change21_22",
           into = "houseprice_change21_22",
           sep = "%") %>% 
  mutate(houseprice_change21_22 = as.numeric(houseprice_change21_22))
```

@@

```{r join population, unemployment, income, and house price data}
# join employment data with income, house prices, tax, and population data
MSApop_income_housecost_tax_emp <- MSA_employment_yoy %>%
  filter(year == "2022") %>% 
  mutate(employment_rate = round(employment_rate, digits = 3)) %>% 
  select(cbsa, year, employment_rate, employment_YoY) %>% 
  # right_join(MSA_property_tax) %>% 
  select(!year)

@@

```{r data table}
# table seperated by MSAs
datatable(MSApop_income_housecost_tax_emp, extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
  )
)
```

@@

```{r MSAs seperate into cities}
#  split up MSAs into cities to collect utility, rent, and landlord data.
msa_seperated <- MSApop_income_housecost_tax_emp %>% 
  separate_rows(MSA, sep = "-") %>% 
  rename(city = MSA)# %>% # seperate MSA cities into their own columns
 # full_join(payroll_data, by = c("MSA" = "area")) %>% 
 # drop_na(pop_percent_change20_21)

>>>>

# create object with viable MSAs 
potential_MSAs <- msa_pop_2021_data %>% 
  filter(pop_percent_change > 0.5 ) %>% 
  mutate(pop_percent_change20_21 = round(pop_percent_change, digits = 3)) %>% 
  filter (!state %in% c("CA", "HI", "AK")) %>% 
  #separate(col = "state", into = c("state1", "state2", "state3"),    sep = "-") %>% 
  filter(!cbsa %in% c("10580", "10900", "12420", "12940", "14260", "15980", "18580", "19100", "19660", "19740", "23540", "25420", "26420", "27260", "29820", "34940", "36260", "36740", "38060", "39100", "39300", "39340", "41620", "45220", "45300", "46060", "46140", "39900")) %>%  # remove MSAs as instructed by Owen 
  mutate(normalized_pop2021 = (popestimate2021-min(popestimate2021))/ (max(popestimate2021)-min(popestimate2021))) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021, popestimate2021)

@@

# join population growth data set to household median income
income_potential_MSAs <- MSA_income_data %>% # rename row names
  select(cbsa, median_income) %>% 
  inner_join(potential_MSAs) %>% 
  select(cbsa, MSA, state, median_income, pop_percent_change20_21, popestimate2021, normalized_pop2021) #reorder columns

@@

# data wrangling, only selecting data for potential MSAs
MSA_employment <- unemployment_data %>% 
  filter(year >= 2019) %>% 
  full_join(income_potential_MSAs) %>% 
  drop_na() %>% 
  select(cbsa, MSA, year, month, employment_rate)

avg_emp_yoy <- MSA_employment %>% 
  group_by(MSA, year) %>% 
  summarize(avg_employment_rate = mean(employment_rate)) %>% 
  mutate(avg_emp_YoY = (avg_employment_rate-lag(avg_employment_rate))/lag(avg_employment_rate)*100) %>% 
  mutate(avg_emp_YoY = round(avg_emp_YoY, digits = 3)) %>% 
  drop_na()

MSA_emp_income <- income_potential_MSAs %>% 
  right_join(avg_emp_yoy) %>% 
  filter(year == "2021") %>% 
  select(!c(avg_employment_rate, year))

@@

@@

```{r employment visuals, fig.align = "center", fig.show="hold", fig.height= 7, fig.width=10, eval = FALSE}
# average employment from 2000 to 2022
employment_summary_since2000 <- MSA_employment %>% 
  group_by(MSA) %>% 
  filter(year >= 2000) %>% 
  summarize(
    avg_employment = mean(employment_rate),
    sd_employment = sd(employment_rate)
  )

# visual of MSA employment rates, shows major drop during covid
ggplot(MSA_employment_yoy, aes(x = year, y = employment_rate, color = MSA)) +
  geom_line() +
  xlim(2021, 2022) +
  theme(legend.position = "none") +

@@

# year over year change in employment rate, split up by year
ggplot(MSA_employment_yoy, aes(x = employment_YoY, y = reorder(MSA, employment_YoY), fill = emp_pos)) +
  geom_col(position = "identity", color = "black", size = 0.25) +
  scale_fill_manual(values = c("red", "darkolivegreen3"), guide = FALSE) +
  facet_grid(~year) +
  labs(x = "Year-over_year change in employment rate",
       y = " ")

@@

```{r payroll visual, fig.height = 7, eval = FALSE}
ggplot(MSA_payroll, aes(x = payroll_change, y = reorder(MSA, payroll_change), fill = payroll_pos)) +
  geom_col(position = "identity", color = "black", size = 0.25) +
  scale_fill_manual(values = c("red", "darkolivegreen3"), guide = FALSE) +
  labs(x = "percent change",
       y = "",
       title = "Quarterly change in employee payroll from 2021 to 2022") +
  facet_grid(~year)

@@

MSA_median_rent2022_data <- read_csv(here("data", "raw", "2022MSA_median_rent.csv")) %>% 
  separate(col = cbsasub22,
           into = c("metro", "cbsa"),
           sep = 5) %>% 
  separate(col = cbsa,
           into = c("unknown", "cbsa"),
           sep = -5) %>% 
  mutate(cbsa = as.double(cbsa)) %>% 
  mutate(median_rent2022 = as.numeric(rent_50_0)) %>% 
  select(cbsa, median_rent2022) %>% 
  right_join(MSA_median_rent2021_data)

potential_MSAs <- MSA_median_rent2022_data %>% 
  left_join(MSA_emp_income)

@@

```{r house price data}
metro_house_cost_data <- read_csv(here("data","raw", "metro_home_costs", "metro-home-prices-2022-11-10.csv")) %>% 
  row_to_names(3) %>% 
  clean_names() %>% 
  separate(col = "metropolitan_area",
           into = c("cbsa", 'metro_area'),
           sep = 5) %>% 
  mutate(cbsa = as.numeric(cbsa)) %>% 
  mutate(x2021 = as.numeric(x2021)) %>% 
  mutate(x2021 = x2021 * 1000) %>% 
  rename(house_price_2021 = x2021)

metro_house_prices <- metro_house_cost_data %>%
  select(cbsa, house_price_2021) %>% 
  right_join(potential_MSAs) %>% 
  select(cbsa, MSA, state, pop_percent_change20_21, normalized_pop2021, popestimate2021, median_income, avg_emp_YoY, median_rent2021, median_rent2022, house_price_2021)
```

@@

@@

```{r data table}
# table seperated by MSAs
datatable(metro_house_prices, extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel', 'pdf', 'print')
  )
)
```

@@

```{r MSAs seperate into cities}
#  split up MSAs into cities to collect utility, rent, and landlord data.
msa_seperated <- metro_house_prices %>% 
  separate_rows(MSA, sep = "-") %>% 
  rename(city = MSA)# %>% # seperate MSA cities into their own columns
 # full_join(payroll_data, by = c("MSA" = "area")) %>% 
 # drop_na(pop_percent_change20_21)
